Notes on the Godunov Step in {\tt MAESTRO}

### 2008-04-15 docs/Godunov_notes/paper.tex
fixed a few typos

For spherical problems, we first construct $\Ub$ by mapping $w_0$ onto a Cartesian array and adding it to $\Ubt$.  This is a spherical generalization of equation (\ref{Total Radial Velocity}).  We also map $\partial w_0/\partial r$ onto a Cartesian array.  This is a spherical generalization of equation (\ref{Radial Derivative of w_0}).  Then we use an extrapolation analogous to equations equations (\ref{Radial utrans_L}) and (\ref{Radial utrans_R}).  For example, for x-faces:
\begin{eqnarray}
\ut_L &=& \ut_{\ib-\eb_x}^n + \left(\half - \frac{\dt}{h}u_{\ib-\eb_x}^n\right)\Delta_x\ut_{\ib-\eb_x}^n + \frac{\dt}{2}\left[(\Ubt^n\cdot\eb_r)\left(\frac{\partial w_0}{\partial r}\right)(\eb_r\cdot\eb_x)\right]_{\ib-\eb_x}, \\
\ut_R &=& \ut_{\ib}^n + \left(\half - \frac{\dt}{h}u_{\ib}^n\right)\Delta_x\ut_{\ib}^n + \frac{\dt}{2}\left[(\Ubt^n\cdot\eb_r)\left(\frac{\partial w_0}{\partial r}\right)(\eb_r\cdot\eb_x)\right]_{\ib}.
\end{eqnarray}
Then, we use the same Riemann solver as the Cartesian case.  Finally, we call {\tt addw0} to convert $\Ubt^{\trans}$ to $\Ub^{\trans}$.  To compute $\vt^{\trans}$ or $\wt^{\trans}$, simply replace all instaces of $\ut$ with $\vt$ (or $\wt$) and all instaces of $x$ with $y$ (or $z$).
\section{Computing $\Ub^{\mac}$}
In {\tt advance\_premac.f90}, we call {\tt make\_edge\_state} to compute $\Ubt^{\mac}$.
\subsection{2D Cartesian Case}
Here are the steps to create $u^{\mac}$ for 2D Cartesian problems.  First, we need to compute a time-centered $u$ at transverse faces:
\begin{eqnarray}
u_L &=& u_{\ib-\eb_y}^n + \left(\half - \frac{\dt}{2h}v_{\ib-\eb_y}^n\right)\Delta_y u_{\ib-\eb_y}^n, \\
u_R &=& u_{\ib} - \left(\half + \frac{\dt}{2h}v_{\ib}^n\right)\Delta_y u_{\ib}^n.
\end{eqnarray}
Then we solve a Riemann problem:
\begin{equation}
u_{\ib-\half\eb_y} =
\begin{cases}
\half(u_L + u_R), & \left|v^{\trans}_{\ib-\half\eb_y}\right| < \epsilon \\
u_L, & v^{\trans}_{\ib-\half\eb_y} > 0, \\
u_R, & v^{\trans}_{\ib-\half\eb_y} < 0. \\
\end{cases}
\end{equation}
Then, we extrapolate the normal velocity to normal faces:
\begin{eqnarray}
u_L^{\edge} &=& u_{\ib-\eb_x}^n + \left(\half - \frac{\dt}{2h}u_{\ib-\eb_x}\right)\Delta_x u_{\ib-\eb_x}^n + \frac{\dt}{2}f_{\ib-\eb_x} \nonumber \\
&& - \frac{\dt}{4h}\left(v_{\ib-\eb_x+\half\eb_y}^{\trans}+v_{\ib-\eb_x-\half\eb_y}^{\trans}\right)\left(u_{\ib-\eb_x+\half\eb_y}+u_{\ib-\eb_x-\half\eb_y}\right), \\
u_R^{\edge} &=& u_{\ib}^n + \left(\half - \frac{\dt}{2h}u_{\ib}\right)\Delta_x u_{\ib}^n + \frac{\dt}{2}f_{\ib} \nonumber \\
&& - \frac{\dt}{4h}\left(v_{\ib+\half\eb_y}^{\trans}+v_{\ib-\half\eb_y}^{\trans}\right)\left(u_{\ib+\half\eb_y}+u_{\ib-\half\eb_y}\right). \\
\end{eqnarray}
Finally, we solve a Riemann problem:
\begin{equation}
u_{\ib-\half\eb_x}^{\mac} =
\begin{cases}
0, & \left(u_L^{\edge} \le 0 ~ {\rm AND} ~ u_R^{\edge} \ge 0\right) ~ {\rm OR} ~ \left|u_L^{\edge} + u_R^{\edge}\right| < \epsilon, \\
u_L^{\edge}, & u_L^{\edge} + u_R^{\edge} > 0, \\
u_R^{\edge}, & u_L^{\edge} + u_R^{\edge} < 0. 
\end{cases}
\end{equation}

To compute $\vt^{\mac}$, we use an analogous procedure for computing $u^{\mac}$, except that we need to incorporate the terms in equation (\ref{Perturbational Velocity Equation}) proportional to $\partial w_0/\partial r$.  First, we need to compute a time-centered $\vt$ at transverse faces:
\begin{eqnarray}
\vt_L &=& \vt_{\ib-\eb_x}^n + \left(\half - \frac{\dt}{2h}u_{\ib-\eb_x}^n\right)\Delta_x \vt_{\ib-\eb_x}^n, \\
\vt_R &=& \vt_{\ib} - \left(\half + \frac{\dt}{2h}u_{\ib}^n\right)\Delta_x \vt_{\ib}^n.
\end{eqnarray}
Then we solve a Riemann problem:
\begin{equation}
\vt_{\ib-\half\eb_x} =
\begin{cases}
\half(\vt_L + \vt_R), & \left|u^{\trans}_{\ib-\half\eb_x}\right| < \epsilon \\
\vt_L, & u^{\trans}_{\ib-\half\eb_x} > 0, \\
\vt_R, & u^{\trans}_{\ib-\half\eb_x} < 0. \\
\end{cases}
\end{equation}
Then, we extrapolate the normal velocity to normal faces, incorporating the extra term proportional to $\partial w_0/\partial r$\footnote{The underbraced terms could be replaced by the average of the neighboring $v^{\trans}$}:
\begin{eqnarray}
\vt_L^{\edge} &=& \vt_{\ib-\eb_y}^n + \left(\half - \frac{\dt}{2h}v_{\ib-\eb_y}\right)\Delta_y \vt_{\ib-\eb_y}^n + \frac{\dt}{2}f_{\ib-\eb_y} \nonumber \\
&& - \frac{\dt}{4h}\left(u_{\ib+\half\eb_x-\eb_y}^{\trans}+u_{\ib-\half\eb_x-\eb_y}^{\trans}\right)\left(\vt_{\ib+\half\eb_x-\eb_y}+\vt_{\ib-\half\eb_x-\eb_y}\right) \nonumber \\
&& - \frac{\dt}{2h}\underbrace{\vt_{\ib-\eb_y}}_{?}(w_{0,j-\half} - w_{0,j-\frac{3}{2}}), \\
\vt_R^{\edge} &=& \vt_{\ib}^n + \left(\half - \frac{\dt}{2h}v_{\ib}\right)\Delta_y \vt_{\ib}^n + \frac{\dt}{2}f_{\ib} \nonumber \\
&& - \frac{\dt}{4h}\left(u_{\ib+\half\eb_x}^{\trans}+u_{\ib-\half\eb_x}^{\trans}\right)\left(\vt_{\ib+\half\eb_x}+\vt_{\ib-\half\eb_x}\right) \nonumber \\
&& - \frac{\dt}{2h}\underbrace{\vt_{\ib}}_{?}(w_{0,j+\half} - w_{0,j-\half}).
\end{eqnarray}
Finally, we solve a Riemann problem:
\begin{equation}
\vt_{\ib-\half\eb_y}^{\mac} =
\begin{cases}
0, & \left(v_L^{\edge} \le 0 ~ {\rm AND} ~ v_R^{\edge} \ge 0\right) ~ {\rm OR} ~ \left|v_L^{\edge} + v_R^{\edge}\right| < \epsilon, \\
\vt_L^{\edge}, & v_L^{\edge} + v_R^{\edge} > 0, \\
\vt_R^{\edge}, & v_L^{\edge} + v_R^{\edge} < 0. 
\end{cases}
\end{equation}

>>>>

For spherical problems, we first construct $\Ub$ by mapping $w_0$ onto a Cartesian array and adding it to $\Ubt$.  This is a spherical generalization of equation (\ref{Total Radial Velocity}).  We also map $\partial w_0/\partial r$ onto a Cartesian array.  This is a spherical generalization of equation (\ref{Radial Derivative of w_0}).  Then we use an extrapolation analogous to equations equations (\ref{Radial utrans_L}) and (\ref{Radial utrans_R}).  For example, for x-faces:
\begin{eqnarray}
\ut_L &=& \ut_{\ib-\eb_x}^n + \left(\half - \frac{\dt}{h}u_{\ib-\eb_x}^n\right)\Delta_x\ut_{\ib-\eb_x}^n + \frac{\dt}{2}\left[(\Ubt^n\cdot\eb_r)\left(\frac{\partial w_0}{\partial r}\right)(\eb_r\cdot\eb_x)\right]_{\ib-\eb_x}, \\
\ut_R &=& \ut_{\ib}^n + \left(\half - \frac{\dt}{h}u_{\ib}^n\right)\Delta_x\ut_{\ib}^n + \frac{\dt}{2}\left[(\Ubt^n\cdot\eb_r)\left(\frac{\partial w_0}{\partial r}\right)(\eb_r\cdot\eb_x)\right]_{\ib}.
\end{eqnarray}
Then, we use the same Riemann solver as the Cartesian case to obtain $\ut^{\trans}$.  To compute $\vt^{\trans}$ or $\wt^{\trans}$, simply replace all instaces of $\ut$ with $\vt$ (or $\wt$) and all instaces of $x$ with $y$ (or $z$).  Finally, we call {\tt addw0} to convert $\Ubt^{\trans}$ to $\Ub^{\trans}$.
\section{Computing $\Ub^{\mac}$}
In {\tt advance\_premac.f90}, we call {\tt make\_edge\_state} to compute $\Ubt^{\mac}$.
\subsection{2D Cartesian Case}
Here are the steps to create $u^{\mac}$ for 2D Cartesian problems.  First, we need to compute a time-centered $u$ at transverse faces:
\begin{eqnarray}
u_L &=& u_{\ib-\eb_y}^n + \left(\half - \frac{\dt}{2h}v_{\ib-\eb_y}^n\right)\Delta_y u_{\ib-\eb_y}^n, \\
u_R &=& u_{\ib} - \left(\half + \frac{\dt}{2h}v_{\ib}^n\right)\Delta_y u_{\ib}^n.
\end{eqnarray}
Then we solve a Riemann problem:
\begin{equation}
u_{\ib-\half\eb_y} =
\begin{cases}
\half(u_L + u_R), & \left|v^{\trans}_{\ib-\half\eb_y}\right| < \epsilon \\
u_L, & v^{\trans}_{\ib-\half\eb_y} > 0, \\
u_R, & v^{\trans}_{\ib-\half\eb_y} < 0. \\
\end{cases}
\end{equation}
Then, we extrapolate the normal velocity to normal faces:
\begin{eqnarray}
u_L^{\edge} &=& u_{\ib-\eb_x}^n + \left(\half - \frac{\dt}{2h}u_{\ib-\eb_x}\right)\Delta_x u_{\ib-\eb_x}^n + \frac{\dt}{2}f_{\ib-\eb_x} \nonumber \\
&& - \frac{\dt}{4h}\left(v_{\ib-\eb_x+\half\eb_y}^{\trans}+v_{\ib-\eb_x-\half\eb_y}^{\trans}\right)\left(u_{\ib-\eb_x+\half\eb_y}+u_{\ib-\eb_x-\half\eb_y}\right), \\
u_R^{\edge} &=& u_{\ib}^n + \left(\half - \frac{\dt}{2h}u_{\ib}\right)\Delta_x u_{\ib}^n + \frac{\dt}{2}f_{\ib} \nonumber \\
&& - \frac{\dt}{4h}\left(v_{\ib+\half\eb_y}^{\trans}+v_{\ib-\half\eb_y}^{\trans}\right)\left(u_{\ib+\half\eb_y}+u_{\ib-\half\eb_y}\right). \\
\end{eqnarray}
Finally, we solve a Riemann problem:
\begin{equation}
u_{\ib-\half\eb_x}^{\mac} =
\begin{cases}
0, & \left(u_L^{\edge} \le 0 ~ {\rm AND} ~ u_R^{\edge} \ge 0\right) ~ {\rm OR} ~ \left|u_L^{\edge} + u_R^{\edge}\right| < \epsilon, \\
u_L^{\edge}, & u_L^{\edge} + u_R^{\edge} > 0, \\
u_R^{\edge}, & u_L^{\edge} + u_R^{\edge} < 0. 
\end{cases}
\end{equation}

To compute $\vt^{\mac}$, we use an analogous procedure for computing $u^{\mac}$, except that we need to incorporate the terms in equation (\ref{Perturbational Velocity Equation}) proportional to $\partial w_0/\partial r$.  First, we need to compute a time-centered $\vt$ at transverse faces:
\begin{eqnarray}
\vt_L &=& \vt_{\ib-\eb_x}^n + \left(\half - \frac{\dt}{2h}u_{\ib-\eb_x}^n\right)\Delta_x \vt_{\ib-\eb_x}^n, \\
\vt_R &=& \vt_{\ib} - \left(\half + \frac{\dt}{2h}u_{\ib}^n\right)\Delta_x \vt_{\ib}^n.
\end{eqnarray}
Then we solve a Riemann problem:
\begin{equation}
\vt_{\ib-\half\eb_x} =
\begin{cases}
\half(\vt_L + \vt_R), & \left|u^{\trans}_{\ib-\half\eb_x}\right| < \epsilon \\
\vt_L, & u^{\trans}_{\ib-\half\eb_x} > 0, \\
\vt_R, & u^{\trans}_{\ib-\half\eb_x} < 0. \\
\end{cases}
\end{equation}
Then, we extrapolate the normal velocity to normal faces, incorporating the extra term proportional to $\partial w_0/\partial r$\footnote{The underbraced terms could be replaced by the average of the neighboring $v^{\trans}$.  We still don't know what the best option is.}:
\begin{eqnarray}
\vt_L^{\edge} &=& \vt_{\ib-\eb_y}^n + \left(\half - \frac{\dt}{2h}v_{\ib-\eb_y}\right)\Delta_y \vt_{\ib-\eb_y}^n + \frac{\dt}{2}f_{\ib-\eb_y} \nonumber \\
&& - \frac{\dt}{4h}\left(u_{\ib+\half\eb_x-\eb_y}^{\trans}+u_{\ib-\half\eb_x-\eb_y}^{\trans}\right)\left(\vt_{\ib+\half\eb_x-\eb_y}+\vt_{\ib-\half\eb_x-\eb_y}\right) \nonumber \\
&& - \frac{\dt}{2h}\underbrace{\vt_{\ib-\eb_y}}_{?}(w_{0,j-\half} - w_{0,j-\frac{3}{2}}), \\
\vt_R^{\edge} &=& \vt_{\ib}^n + \left(\half - \frac{\dt}{2h}v_{\ib}\right)\Delta_y \vt_{\ib}^n + \frac{\dt}{2}f_{\ib} \nonumber \\
&& - \frac{\dt}{4h}\left(u_{\ib+\half\eb_x}^{\trans}+u_{\ib-\half\eb_x}^{\trans}\right)\left(\vt_{\ib+\half\eb_x}+\vt_{\ib-\half\eb_x}\right) \nonumber \\
&& - \frac{\dt}{2h}\underbrace{\vt_{\ib}}_{?}(w_{0,j+\half} - w_{0,j-\half}).
\end{eqnarray}
Finally, we solve a Riemann problem:
\begin{equation}
\vt_{\ib-\half\eb_y}^{\mac} =
\begin{cases}
0, & \left(v_L^{\edge} \le 0 ~ {\rm AND} ~ v_R^{\edge} \ge 0\right) ~ {\rm OR} ~ \left|v_L^{\edge} + v_R^{\edge}\right| < \epsilon, \\
\vt_L^{\edge}, & v_L^{\edge} + v_R^{\edge} > 0, \\
\vt_R^{\edge}, & v_L^{\edge} + v_R^{\edge} < 0. 
\end{cases}
\end{equation}
